---
title: "Assesing Demand for Transit Accessesible Apartments in Chicago"
author: "Noah Anderson"
output: html_document
---

```{r echo=True, include=FALSE}

library(tidyverse)
library(tidycensus)
library(sf)
library(kableExtra)
library(acs)

options(scipen=999)
options(tigris_class = "sf")

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

palette5 <- c("#f0f9e8","#bae4bc","#7bccc4","#43a2ca","#0868ac")
```


# Introduction
This project uses American Community Studies data to assess demand for rental units with transit access in the City of Chicago. The goal is to Transit Oriented Development (TOD) 

https://www.chicago.gov/city/en/sites/equitable-transit-oriented-development/home.html#:~:text=ETOD%20Pre%2DDevelopment%20Grants%20Up%20to%20%24150%2C000!&text=Apply%20by%20August%2018%2C%202023.,-Learn%20More&text=Every%20Chicagoan%20should%20be%20able,%2DOriented%20Development%20(ETOD).

```{r echo=True, include=FALSE}

cook09 <- 
  get_acs(geography = "tract", variables = c("B25026_001E","B02001_002E","B15001_050E",
                                             "B15001_009E","B19013_001E","B25058_001E",
                                             "B06012_002E"), 
          year=2009, state=17, county=031, geometry=T, output="wide", cache_table = TRUE) %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, Whites = B02001_002E,
         FemaleBachelors = B15001_050E, MaleBachelors = B15001_009E,
         MedHHInc = B19013_001E, MedRent = B25058_001E, 
         TotalPoverty = B06012_002E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         year = "2009", county = "Cook") %>%
  dplyr::select(-Whites, -FemaleBachelors, -MaleBachelors, -TotalPoverty) 


cook19 <- 
  get_acs(geography = "tract", variables = c("B25026_001E","B02001_002E","B15001_050E",
                                             "B15001_009E","B19013_001E","B25058_001E",
                                             "B06012_002E"), 
          year=2019, state=17, county=031, geometry=T, output="wide", cache_table = TRUE) %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, Whites = B02001_002E,
         FemaleBachelors = B15001_050E, MaleBachelors = B15001_009E,
         MedHHInc = B19013_001E, MedRent = B25058_001E, 
         TotalPoverty = B06012_002E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         year = "2019", county = "Cook") %>%
  dplyr::select(-Whites, -FemaleBachelors, -MaleBachelors, -TotalPoverty) 

```






```{r}
chicago_neighborhoods <- st_read("https://raw.githubusercontent.com/noahba65/chicago-TOD/main/DATA/chicago_boundary.geojson") %>%
  st_transform('ESRI:102728') %>% st_sf()



chicago_boundary <- st_union(chicago_neighborhoods) %>% 
  st_sf()

cta_stops <- st_read("https://raw.githubusercontent.com/ChicagoCityscape/gis-data/master/stations_cta/cta_entrances.geojson") %>%
  st_transform('ESRI:102728') %>% st_sf()

# Creating Chicago only ACS tract data for 2009 and 2019
chicago09 <- st_intersection(chicago_boundary, cook09)
chicago19 <- st_intersection(chicago_boundary,cook19) 

allTracts_chicago <- rbind(chicago09,chicago19)


# Creating CTA buffer
cta_buffer <- st_union(st_buffer(cta_stops,5280/4)) %>%
  st_sf()



tract_buffer_chicago <- st_centroid(chicago09)[cta_buffer,] %>%
  st_drop_geometry() %>%
  left_join(dplyr::select(chicago09,GEOID)) %>%
  st_sf()


ggplot() +
  geom_sf(data = st_union(chicago09)) +
  geom_sf(data = tract_buffer_chicago, fill = "red")  + mapTheme()

```



```{r}

# Combining CMAP 2009 and 2019 ACS data 
allTracts_chicago.group <- 
  rbind( 
    st_centroid(allTracts_chicago)[tract_buffer_chicago,] %>%
      st_drop_geometry() %>%
      left_join(allTracts_chicago) %>%
      st_sf() %>%
      mutate(TOD = "TOD"),
    st_centroid(allTracts_chicago)[tract_buffer_chicago, op = st_disjoint] %>%
      st_drop_geometry() %>%
      left_join(allTracts_chicago) %>%
      st_sf() %>%
      mutate(TOD = "Non-TOD")) %>%
  mutate(MedRent.inf = ifelse(year == "2009", MedRent * 1.2, MedRent)) 
# 2009 rent is adjusted for inflation 

allTracts_chicago.Summary <- 
  st_drop_geometry(allTracts_chicago.group) %>%
    group_by(year, TOD) %>%
    summarize(Rent = mean(MedRent.inf, na.rm = T),
              Population = mean(TotalPop, na.rm = T),
              Percent_White = mean(pctWhite, na.rm = T),
              Percent_Bach = mean(pctBachelors, na.rm = T),
              Percent_Poverty = mean(pctPoverty, na.rm = T))

kable(allTracts_chicago.Summary) %>%
  kable_styling() %>%
  footnote(general_title = "\n",
           general = "Table 1.2")


allTracts_chicago.Summary %>%
  gather(Variables, Value, -year, -TOD) %>%
  ggplot(aes(year, Value, fill = TOD)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~Variables, scales = "free", ncol=5) +
    scale_fill_manual(values = c("#bae4bc", "#0868ac")) +
    labs(title = "Indicator differences across time and space")  + theme(legend.position="bottom")



```
Rent as a function of distance to stations CMAP
```{r}
allTracts_chicago.rings <-
  st_join(st_centroid(dplyr::select(chicago19, GEOID, year)), 
          multipleRingBuffer(st_union(cta_stops), 47520, 5280/2)) %>%
  st_drop_geometry() %>%
  left_join(dplyr::select(allTracts_chicago, GEOID, MedRent, year), 
            by=c("GEOID"="GEOID", "year"="year")) %>%
  st_sf() %>%
  mutate(distance = distance / 5280) #convert to miles

medRent_Distance_chicago <-  allTracts_chicago.rings %>%
  group_by(distance) %>%
  summarize(mean_rent = mean(MedRent, na.rm = TRUE))


ggplot() + 
  
  geom_line(data = medRent_Distance_chicago, aes(x = distance, y = mean_rent)) +
  xlab("Distance (Miles)") + ylab("Mean Rent (USD)") + ggtitle("Rent as a Function of Distance to L Stops")



```

Looking at a Rent map, we can see that the area near Downtown has much higher rents. This could be the cause of some of the high value associated with TOD regions. 
```{r}
ggplot() +
  geom_sf(data = allTracts_chicago, aes(fill = MedRent)) + 
  labs(fill = "Median Rent (USD)") + ggtitle("Rent Distribution in Chicago") + 
  mapTheme() 
```


Select neighborhoods in downtown and exclude them. 

```{r}
downtown_neighborhoods <- chicago_neighborhoods %>%
  filter(pri_neigh == "Loop" | pri_neigh == "West Loop" | pri_neigh == "River North" | pri_neigh == "Gold Coast" | pri_neigh == "Streeterville" | pri_neigh == "Lakeshore East" | pri_neigh == "South Loop"  )

downtown_boundary <- st_union(downtown_neighborhoods)

# Excluded neighborhhods
ggplot() +
  geom_sf(data = allTracts_chicago)+
  geom_sf(data = downtown_boundary, fill = "red")  + 
  
 
   mapTheme()


no_downtown <- st_difference(allTracts_chicago, downtown_boundary)

no_downtown.group <- st_difference(allTracts_chicago.group,downtown_boundary)
```

now repeat summary steps but excluding Chicago. 
```{r}
no_downtown.Summary <- 
  st_drop_geometry(no_downtown.group) %>%
    group_by(year, TOD) %>%
    summarize(Rent = mean(MedRent.inf, na.rm = T),
              Population = mean(TotalPop, na.rm = T),
              Percent_White = mean(pctWhite, na.rm = T),
              Percent_Bach = mean(pctBachelors, na.rm = T),
              Percent_Poverty = mean(pctPoverty, na.rm = T))

kable(no_downtown.Summary) %>%
  kable_styling() %>%
  footnote(general_title = "\n",
           general = "Table 1.2")


no_downtown.Summary %>%
  gather(Variables, Value, -year, -TOD) %>%
  ggplot(aes(year, Value, fill = TOD)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~Variables, scales = "free", ncol=5) +
    scale_fill_manual(values = c("#bae4bc", "#0868ac")) +
    labs(title = "Indicator differences across time and space")  + theme(legend.position="bottom")



```


```{r}


no_downtown.rings <-
  st_join(st_centroid(dplyr::select(no_downtown, GEOID, year)), 
          multipleRingBuffer(st_union(cta_stops), 47520, 5280/2)) %>%
  st_drop_geometry() %>%
  left_join(dplyr::select(no_downtown, GEOID, MedRent, year), 
            by=c("GEOID"="GEOID", "year"="year")) %>%
  st_sf() %>%
  mutate(distance = distance / 5280) #convert to miles

medRent_Distance_no_downtown <-  no_downtown.rings %>%
  group_by(distance) %>%
  summarize(mean_rent = mean(MedRent, na.rm = TRUE))


ggplot() + 
  
  geom_line(data = medRent_Distance_no_downtown, aes(x = distance, y = mean_rent)) +
  xlab("Distance (Miles)") + ylab("Mean Rent (USD)") + ggtitle("Rent as a Function of Distance to L Stops")



```
